Refining character relationships using embeddings of textual units

### 2022-09-06 article/paper1.tex
write article

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
\usepackage{bm}
%% auto break lines
\lstset{breaklines=true}

@@

The second table, denoted by $\mathbf{O}$, has a size of $(n \times p)$ where $p$ is the number of \emph{character-based objects} found in the narrative and cells $o_{ij}$ counts occurrences of object $j$ in unit $i$. A character-based object can be loosely defined in order to be flexible for various types of narrative or analyses, but can be roughly seen as a countable recurring narrative entity, containing one or more characters. For example, it can be a character occurrence, a character co-occurrence, an oriented character interaction (e.g. a dialog), or even a particular recurring event containing multiple characters (e.g. a lecture). In this article, we mostly consider character occurrences and pair of characters co-occurrences, as shown in Table \ref{occ_table}.

@@

This data organization already gives an orientation to the subsequent analyses and should be kept in mind of the practitioner. Textual unit are now considered as \emph{individuals} (in the statistical terminology), defined by their \emph{variables} contained in the different columns of both tables. Moreover, subsequent analyses are oriented in searching how the object table $\mathbf{O}$ has an influence over the contingency table $\mathbf{N}$, i.e. searching which words are over-represented or under-represented knowing the character-based objects in the unit. While authors use characters in order to build the narrative, we, to a certain extent, work backward: we are searching how character appearances and interactions in the textual unit act on her/his choice of words. If extraction methods permit it, a practitioner should include all character-based objects which she/he desire to study. Here, for example, the choice to include character co-occurrences along with occurrences is motivated by the fact that we are interested in studying character relationships. A character co-occurrence can roughly be seen as an interaction between two characters, and an interaction between two characters (or more, but higher order interactions are outside the scope of this article) should be considered as an object of its own: this interaction is not necessarily the sum of its parts and gives a particular flavor to the unit.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it, such as Sentiment Analysis (REF), Textometry (REF), or even Deep Learning methods. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. \\
In section (REF), these vectors of textual units are used as anchor points in order to also embed character-based objects in the same space. Therefore, it is crucial that an interpretation about the directions, or the regions, of this embedding is possible, in order to properly extract  information about the localization of character-based objects vectors (the relative position of these vectors is insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of character-based objects relatively to word vectors, these objects can be characterized.

@@

Using Correspondence Analysis in order to analyze textual resources has a long tradition in literature (REF). It has the advantage to naturally provide an embedding space, the factorial map, where units are placed alongside word vectors, allowing to analyze or represent the different units in terms of word specificities. Moreover, units and words placement in the embedding space have a direct interpretation in terms of chi2 distance between profiles, which is desirable when interpreting results.

By performing a Correspondence Analysis on table $\mathbf{N}$, we get $n$ vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ corresponding to units (rows) and $v$ vectors $\mathbf{w}_1, \ldots, \mathbf{w}_v$ corresponding to words (columns). Each of these vectors has a size of $\min(n, v) - 1$, which will generally be $n - 1$. For detailed computations of quantities given by the CA, see Appendix \ref{ca_details}. The position of a unit vector $\mathbf{x}_i$ can be interpreted by two different methods.

The first methods is to look at its position on a particular axis $\alpha$, i.e. looking at the component $x_{i\alpha}$. This component gives the position of the unit on a particular found "contrast" between units, depending on the chosen $\alpha$, and this contrast will be strong if $p_\alpha$, i.e. \emph{the proportion of inertia expressed in $\alpha$}, is high. It can further be interpreted by observing the \emph{contribution of each word $j$} composing the axis $\alpha$, i.e. $c^w_{j\alpha}$ and looking at the sign of $w_{j\alpha}$. Hopefully, by extracting and interpreting the most positive and negative contributing words, the chosen axis can express a particular duality which operates inside the studied text (e.g. love-hatred, action-description, lightness-darkness). The second method is to directly look at the \emph{similarities} $s_{ij}$ between an unit $i$ and a word $j$, as defined by the scalar product between their vectors $s_{ij} := \mathbf{x}^\top_i \mathbf{w}_j$. A positive (resp. negative) similarity denotes a over-representation (resp. under-representation) of the word $j$ in $i$, which permit to find lists of words characterizing the different units. Observe that in this article, both methods are rather applied to character-based object vectors defined in section (REF), as they lay in the same space as unit vectors.

Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from CA reflect textual unit profile (in terms of words) regarding the mean profile. This mean that this analysis is in fact \emph{contrastive}: we highlight unit variations among the studied text. This could therefore becomes problematic if we would like to build our dataset on multiple works: units belonging to the same text have a greater chance to be close from each other, and the displayed variance might be largely composed by differences of style between text. Moreover, the words helping the interpretation of units (and character-based objects) are contained in the studied text, which might limit its range of application.

\subsubsection{Pre-trained word vector}

Pre-trained word vectors, based on Word2Vec (REF), Glove (REF), or Fasttext (REF), have recently received great attention from various fields (REFS). The generally originate from a training of very large corpora, such as Wikipedia (REF) or Common Crawl (REF) and results in a embedding containing a large quantities of word vectors.

\subsubsection{Character and character pairs embeddings}

@@

where $\lambda_\alpha$ are the eigenvalues contained in the diagonal of $\bm{\Lambda}$ and $u_{i \alpha}$ the eigenvectors components found in $\mathbf{U}$. \\
We can also find the vector of word $j$, $\mathbf{w}_j = (w_{j\alpha})$, with

>>>>

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
\usepackage{bm}
\usepackage{url}

%% auto break lines
\lstset{breaklines=true}

@@

The second table, denoted by $\mathbf{O}$, has a size of $(n \times p)$ where $p$ is the number of \emph{character-based objects} found in the narrative and cells $o_{ij}$ counts occurrences of object $j$ in unit $i$. A character-based object can be loosely defined in order to be flexible for various types of narrative or analyses, but can be roughly seen as a countable recurring narrative entity, containing one or more characters. For example, it can be a character occurrence, a character co-occurrence, an oriented character interaction (e.g. a dialog), or even a particular recurring event containing multiple characters (e.g. a meeting). In this article, we mostly consider character occurrences and pair of characters co-occurrences, as shown in Table \ref{occ_table}.

@@

This data organization already gives an orientation to the subsequent analyses and should be kept in mind of the practitioner. Textual unit are now considered as \emph{individuals} (in the statistical terminology), defined by their \emph{variables} contained in the different columns of both tables. Moreover, subsequent analyses are oriented in searching how the object table $\mathbf{O}$ has an influence over the contingency table $\mathbf{N}$, i.e. searching which words are over-represented or under-represented knowing the character-based objects in the unit. While authors use characters in order to build the narrative, we, to a certain extent, work backward: we are searching how character appearances and interactions in the textual unit act on her/his choice of words. If the extraction method permits it, a practitioner should include all character-based objects which she/he desire to study. Here, for example, the choice to include character co-occurrences along with occurrences is motivated by the fact that we are interested in studying character relationships. A character co-occurrence can roughly be seen as an interaction between two characters, and an interaction between two characters (or more, but higher order interactions are outside the scope of this article) should be considered as an object of its own: this interaction is not necessarily the sum of its parts and can give a particular flavor to the unit.

@@

Various methods can be performed on the contingency table $\mathbf{N}$ in order to extract information from it, such as Sentiment Analysis (REF), Textometry (REF), or even Deep Learning methods. Here, we make to choice to extract a lower dimensional, numeric representation of each units, in other words, an \emph{embedding of textual units}. \\
In section (REF), these vectors of textual units are used as anchor points in order to also embed character-based objects in the same space. Therefore, it is crucial that an interpretation about the directions, or the regions, of this embedding is possible in order to properly extract information about the localization of character-based objects vectors (the relative position of these vectors is insufficient). For that reason, we focus on embeddings of textual units which also contain \emph{vectors of words}: by examining the positions of character-based objects relatively to word vectors, these objects can be characterized.

@@

Using Correspondence Analysis in order to analyze textual resources has a long tradition in literature (REF). It has the advantage to naturally provide an embedding space, the factorial map, where units are placed alongside word vectors, allowing to analyze or represent the different units in terms of word frequency profiles. Moreover, units and words placement in the embedding space have a direct interpretation in terms of chi2 distance between profiles, which is desirable when interpreting results.

By performing a Correspondence Analysis on table $\mathbf{N}$, we get $n$ vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ corresponding to units (rows) and $v$ vectors $\mathbf{w}_1, \ldots, \mathbf{w}_v$ corresponding to words (columns). Each of these vectors has a size of $\min(n, v) - 1$, which will generally be $n - 1$. For detailed computations of quantities given by the CA, see Appendix \ref{ca_details}. The position of a unit vector $\mathbf{x}_i$ can be interpreted by mutliple approaches, but we mainly use two here.

The first way to interpret it is to look at the unit-vector position on a particular axis $\alpha$, i.e. looking at the component $x_{i\alpha}$. This component gives the position of the unit on a particular found "contrast" between units, depending on the chosen $\alpha$, and this contrast will be strong if $p_\alpha$, i.e. \emph{the proportion of inertia expressed in $\alpha$}, is high. It can further be interpreted by observing the \emph{contribution of each word $j$} composing the axis $\alpha$, i.e. $c^w_{j\alpha}$ and looking at the sign of $w_{j\alpha}$. Hopefully, by extracting and interpreting the most positive and negative contributing words, the chosen axis can express a particular duality which operates inside the studied text (e.g. love-hatred, action-description, lightness-darkness). The second method is to directly look at the \emph{similarities} $s_{ij}$ between an unit $i$ and a word $j$, as defined by the scalar product between their vectors $s_{ij} := \mathbf{x}^\top_i \mathbf{w}_j$. A positive (resp. negative) similarity denotes a over-representation (resp. under-representation) of the word $j$ in $i$, which permit to find lists of words characterizing the different units. Observe that in this article, both methods are rather applied to character-based object vectors defined in section (REF), as they lay in the same space as unit vectors.

Using CA to embedded units however comes with some limitations. First, note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from CA reflect textual unit profile (in terms of words) regarding the mean profile. This analysis is therefore \emph{contrastive}: it highlights unit variations among the studied text. This could becomes problematic in order to study a dataset containing multiple texts: units belonging to the same text will have a greater chance to be close from each other, and the displayed variance might be largely composed by differences of style between text. Another limitation with this approach is that the words helping the interpretation of units (and character-based objects) are contained in the studied text. Approaches requiring to study the position of units and objects relatively to a predefined list of words (e.g., friends, enemies, family) might therefore be impossible.

\subsubsection{Pre-trained word vector (WV)}

Pre-trained word vectors, based on Word2Vec (REF), Glove (REF), or fastText (REF), have recently received great attention from various fields (REFS). They are generally obtained through a training on a very large corpora, such as Wikipedia (REF) or Common Crawl (REF), and the resulting embedding contains a large quantity of word vectors. As shown by multiple studies (REF), these vectors are placed in order to reflect semantic and syntactic similarities between words, and are used in mutliple applications (REF).

There exists various ways to use these pre-trained word vectors in order to find vectors for a \emph{group of words}, such as sentences (REF), paragraphs (REF), or documents (REF), which are often used to apply a classification or clustering algorithm on the newly embedded objects (REF). Here, we use a recent methodology proposed in (REF), in order to obtain textual unit vectors laying in the same space as words.

More precisely, we use a pre-trained word vectors $w_1, \ldots, w_m$ trained on Common Crawl and Wikipedia using fastText.\footnote{\url{ https://fasttext.cc/docs/en/crawl-vectors.html}, accessed September 2022.} For French, the number of word-vectors $m$ is

\subsection{Character-based object embeddings}

@@

where $\lambda_\alpha$ are the eigenvalues contained in the diagonal matrix $\bm{\Lambda}$ and $u_{i \alpha}$ the eigenvectors components found in $\mathbf{U}$. We find the vector of word $j$, $\mathbf{w}_j = (w_{j\alpha})$, with
